---
output:
  html_document: default
  pdf_document: default
---
# Reproducible Research Project 2 
Minh Tri
04/04/2022

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The basic goal of this report is to explore the NOAA Storm Database by using R and answer some basic questions about severe weather events.  

## Data Processing 
The data for the analysis was downloaded from the NOAA storm database. After the data is downloaded from the website, it is uncompressed and read into R environment
```{r}
setwd("D:/Statistics/R/R data/Reproducible Research Project 2")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "repdata_data_StormData.csv.bz2")
data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"))
```

## What causes the most injuries?
To answer the 1st question, we create a bar chart illustrate the relation between Event type(x axis) and the number of injuries(y axis)

At the beggining, let summerize the number of injuries according to Event type. After that, reorder the event and drawing a bar graph.
```{r}
injuries <- aggregate(data$INJURIES, by = list(EVENT= data$EVTYPE), sum)
injuries <- injuries[order(injuries$x, decreasing = TRUE), ]
head(injuries)
```

```{r}
library(ggplot2)
bar <- ggplot(injuries[1:6, ], aes(EVENT,  x, fill = EVENT, label = x))
bar + stat_summary(geom = "bar") + labs(x = "Event Type", y = "Number of Injuries") + geom_text(nudge_y = 4000) + ggtitle("Most Injuries Events") 
```
Looking at the graph, we can see that TORNADO is responsible for most of injuries (91346 events occurred)

## What causes the most fatalities?
To answer this question, we create a bar chart illustrate the relation between Event type(x axis) and the number of fatalitis(y axis)

At the beggining, let summerize the number of fatalities according to Event type. After that, reorder the event and drawing a bar graph.
```{r}
fatalities <- aggregate(data$FATALITIES, by = list(EVENT= data$EVTYPE), sum)
fatalities <- fatalities[order(fatalities$x, decreasing = TRUE), ]
head(fatalities)
```

```{r}
library(ggplot2)
bar1 <- ggplot(fatalities[1:6, ], aes(EVENT,  x, fill = EVENT, label = x))
bar1 + stat_summary(geom = "bar") + labs(x = "Event Type", y = "Number of fatalities") + geom_text(nudge_y = 200) + ggtitle("Most Fatal Events") 
```
Looking at the graph, we can see that TORNADO is responsible for most of fatalities (5633 events occurred)

## Across the United States, which types of events have the greatest economic consequences?
Checking the all the characters of PRO/CROPDMGEXP variables
```{r}
unique(data$PROPDMGEXP)
```
```{r}
unique(data$CROPDMGEXP)
```


Changing these characters to upper case
```{r}
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)
unique(data$PROPDMGEXP); unique(data$CROPDMGEXP)
```
Assigning numeric value according to the characters: Billion (9), Hundred (2), Kilo (3), and Million (6) 
```{r}
data[data$PROPDMGEXP == "B", "PROPDMGEXP"] <- 9
data[data$PROPDMGEXP == "M", "PROPDMGEXP"] <- 6
data[data$PROPDMGEXP == "K", "PROPDMGEXP"] <- 3
data[data$PROPDMGEXP == "H", "PROPDMGEXP"] <- 2
data[data$PROPDMGEXP %in% c("", "+", "-", "?"), "PROPDMGEXP"] <- "0"
data[data$CROPDMGEXP %in% c("", "+", "-", "?"), "CROPDMGEXP"] <- "0"
data[data$CROPDMGEXP == "B", "CROPDMGEXP"] <- 9
data[data$CROPDMGEXP == "M", "CROPDMGEXP"] <- 6
data[data$CROPDMGEXP == "K", "CROPDMGEXP"] <- 3
data[data$CROPDMGEXP == "H", "CROPDMGEXP"] <- 2
unique(c(data$PROPDMGEXP, data$CROPDMGEXP))
```
Assign the PDMGEXP value 
```{r}
data$PROPDMGEXP <- 10^(as.numeric(data$PROPDMGEXP))
data$CROPDMGEXP <- 10^(as.numeric(data$CROPDMGEXP))
```

Calculate the total damage
```{r}
data$DMGTOTAL <- data$PROPDMGEXP * data$PROPDMG + data$CROPDMGEXP * data$CROPDMG
```

Extract the value (DMGToTAL, EVTYPE)
```{r}
DamageByType  <- aggregate(data$DMGTOTAL, by = list(EVENT= data$EVTYPE), sum)
DamageByType <- DamageByType[order(DamageByType$x, decreasing = TRUE), ]
head(DamageByType)
```

Drawing plot
```{r}
library(ggplot2)
bar2 <- ggplot(DamageByType[1:6, ], aes(EVENT,  x, fill = EVENT))
bar2 + geom_bar(stat = "identity") + labs(x = "Event Type", y = "Total Damage") +  ggtitle("Top 6 events that cause severe damage")  
```
Looking at the graph, we can see that FLOOD is responsible for most of the damage occurred.

# Conclusion
Based on the graph obtained, while FLOOD is responsible for most of the damage, TORNADO is the major cause of human injuries and fatalities. 

